Water desalination using polyelectrolyte hydrogel. Gibbs ensemble modelling

### 2022-07-01 gibbs.tex
theory part

\newcommand{\cna}{c_\mathrm{Na^+}}
\newcommand{\ccl}{c_\mathrm{Cl^-}}
\newcommand{\cca}{c_\mathrm{Ca^{2+}}}
\newcommand{\ch}{c_\mathrm{H^+}}
\newcommand{\cp}{c_\mathrm{p}}
\newcommand{\nna}{N_\mathrm{Na^+}}
\newcommand{\ncl}{N_\mathrm{Cl^-}}
\newcommand{\ncleq}{\widetilde{N}_\mathrm{Cl^-}}
\newcommand{\nca}{N_\mathrm{Ca^{2+}}}

@@

\newcommand{\Ka}{K_{\mathrm{A}}}
\newcommand{\pKa}{\mathrm{p}\Ka}
\newcommand{\pK}{\mathrm{p}K}
\newcommand{\pH}{\mathrm{pH}}
\newcommand{\mol}{\mathrm{mol}}
\newcommand{\molperl}{\mathrm{mol/l}}
\newcommand{\kg}{\mathrm{kg}}
\newcommand{\res}{^{\mathrm{res}}}
\newcommand{\pHres}{\pH\res}
\newcommand{\pHgel}{\pH\gel}
\newcommand{\cs}{c_{\mathrm{s}}}
%\newcommand{\cs}{c_{\mathrm{salt}}}
\newcommand{\csres}{\cs\res}
%\newcommand{\Vgel}{V\gel}
\newcommand{\Vgel}{v_\mathrm{gel}}
\newcommand{\Vgeleq}{\widetilde{\Vgel}}
\newcommand{\Pgel}{P\gel}
\newcommand{\Pres}{P\res}
\newcommand{\Vout}{V\out}
\newcommand{\Vbox}{V\tot}
\newcommand{\PE}{polyelectrolyte{}}

@@

\begin{figure*}[t]
\centering \includegraphics[width=0.5\textwidth]{figures/simgibbs}\caption{Diamond-like network in the simulation box. Color code represents
the individual ion types (red: $\na$, blue: $\cl$, yellow: $\ca$)
and the hydrogel (gray: neutral segment ($\AH$), cyan: charged segment
($\A$)). \label{fig:diamond}.}
\end{figure*}

@@

We model our gel as a network of 16 linear polymer chains, each by
30 monomer units. The chains are connected to a diamond-like polymer
network and put into a simulation box with periodic boundaries. Each
unit of the network carries a negative electric charge, which is equal
to the charge of electron. Except the particles of the network, the
mobile monovalent ions, $\na$ and $\cl$, are present in simulation
box. The total electric charge of all the particles in the box is
zero, therefore the amount of $\na$ ions is bigger than that of $\cl$
by the number of hydrogel units, $N\gel=16\cdot30=480$.

The goal of this study is to simulate the process of water desalination
by means of compression. In order to do it we simulate the gel in
two ensembles.
\begin{enumerate}
\item The first is so called \emph{open system}, when the simulation box
with gel freely exchanges ions with a big amount of aqueous solution
of certain constant salinity. 
\item And in \emph{closed system}, when the gel is in equilibrium with a
finite volume of aqueous solution.
\end{enumerate}
The compression of the gel in the open system does not affect the surrounding salinity, thus the chemical potential of ions remains constant $\muna=$ $\mucl=$ $\text{Const}$. 
On the contrary, what remains constant in the closed system is total number of ions, that are contained in the gel and in the external volume, $\nna=\text{Const}$ and $\ncl=\text{Const}$.
By $\nna$ and $\ncl$ we will understand the density of the corresponding ion in the compression volume
\begin{equation}
\ncl = \left(\ncl\gel + \cs\cdot(\Vbox - \Vgel)\right)/\Vbox
\end{equation}

\begin{figure*}[t]
\centering 
\includegraphics[width=0.5\textwidth]{figures/gc}\includegraphics[width=0.5\textwidth]{figures/gibbs}\caption{Diamond-like network in the simulation box. Color code represents
the individual ion types (red: $\na$, blue: $\cl$, yellow: $\ca$)
and the hydrogel (gray: neutral segment ($\AH$), cyan: charged segment
($\A$)). \label{fig:open and closed}.}
\end{figure*}

@@

The similar scheme we use to simulate the gel in closed system. The
procedure represents a chain of subsequent MD and MC simulation, but
now we run the molecular dynamics in two simulation boxes simultaneously.
The first box of the volume $\Vgel$ contains the gel with ions, and
the second box, $\Vout$--- contains only ions. These two boxes are
represented in Figure~\ref{fig:diamond}. The total volume of both
boxes is kept constant, $V=V_{\mathrm{gel}}+V_{\mathrm{out}}$, so
the compression of the gel implies the decrease $\Vgel$ and increase
$\Vout$.

The thermodynamic equilibrium between the two volumes implies the
exchange of the ion pairs between them, which is done by means of
MC procedure in a way similar to described in \cite{Panagiotopoulos1988b}.

\paragraph{Algorithm.}

In general the simulation algorithm is following:
\begin{enumerate}
\item simulate molecular dynamics in both boxes; collect the observables,
for examples pressure values $P$, to an array; 
\item perform Monte-Carlo procedure simulating ion pair exchange; collect
the arrays of observables sample : number of ions in both boxes, $\nna^{I}$,
$\ncl^{I}$, $\nna^{II}$ and $\ncl^{II}$ ; 
\item repeat the procedure until the desired length of sample arrays is
reached.
\end{enumerate}

>>>>

\newcommand{\cna}{c_\mathrm{Na^+}}
\newcommand{\ccl}{c_\mathrm{Cl^-}}
\newcommand{\cca}{c_\mathrm{Ca^{2+}}}
\newcommand{\ch}{c_\mathrm{H^+}}
\newcommand{\cp}{c_\mathrm{p}}
\newcommand{\nna}{n_\mathrm{Na^+}}
\newcommand{\ncl}{n_\mathrm{Cl^-}}
\newcommand{\ncleq}{\widetilde{N}_\mathrm{Cl^-}}
\newcommand{\nca}{N_\mathrm{Ca^{2+}}}

@@

\newcommand{\Ka}{K_{\mathrm{A}}}
\newcommand{\pKa}{\mathrm{p}\Ka}
\newcommand{\pK}{\mathrm{p}K}
\newcommand{\pH}{\mathrm{pH}}
\newcommand{\mol}{\mathrm{mol}}
\newcommand{\molperl}{\mathrm{mol/l}}
\newcommand{\kg}{\mathrm{kg}}
\newcommand{\res}{^{\mathrm{res}}}
\newcommand{\pHres}{\pH\res}
\newcommand{\pHgel}{\pH\gel}
\newcommand{\cs}{c_{\mathrm{s}}}
%\newcommand{\cs}{c_{\mathrm{salt}}}
\newcommand{\csres}{\cs\res}
%\newcommand{\Vgel}{V\gel}
\newcommand{\Vgel}{V_\mathrm{gel}}
\newcommand{\Ngel}{N_\mathrm{gel}}
\newcommand{\Vgeleq}{\widetilde{\Vgel}}
\newcommand{\Pgel}{P\gel}
\newcommand{\Pres}{P\res}
\newcommand{\Vout}{V_\mathrm{out}}
\newcommand{\Vbox}{V_\mathrm{box}}
\newcommand{\PE}{polyelectrolyte{}}

@@

\begin{figure*}[t]
\centering \includegraphics[width=0.7\textwidth]{figures/simgibbs}\caption{Diamond-like network in the simulation box. Color code represents
the individual ion types (red: $\na$, blue: $\cl$, yellow: $\ca$)
and the hydrogel (gray: neutral segment ($\AH$), cyan: charged segment
($\A$)). \label{fig:diamond}.}
\end{figure*}

@@

We model the gel as a network of 16 linear polymer chains, each by 30 monomer units. 
The chains are connected to a diamond-like polymer network throught 8 crosslinking units, and put into a simulation box with periodic boundaries (see \reffig{fig:diamond}). 
Each unit of the network carries a negative electric charge, which is equal to the charge of electron. 
Except the particles of the network, the monovalent co- and counter-ions ions, $\cl$ and $\na$, are present in simulation box. 
The total electric charge of all the particles in the box is zero, therefore the amount of $\na$ ions is bigger than that of $\cl$ by the number of hydrogel units, $\Ngel=16\cdot30 + 8=488$.

\begin{figure*}[t]
\centering 
\subfloat[Open system \label{fig: open}]
{\includegraphics[width=0.5\textwidth]{figures/gc}}
\subfloat[Closed system \label{fig: closed}]
{\includegraphics[width=0.5\textwidth]{figures/gibbs}}
\caption{Diamond-like network in the simulation box. Color code represents
the individual ion types (red: $\na$, blue: $\cl$, yellow: $\ca$)
and the hydrogel (gray: neutral segment ($\AH$), cyan: charged segment
($\A$)). \label{fig:open and closed}.}
\end{figure*}
We simulate the compression of the gel in two ensembles.
\begin{enumerate}
\item The first is so called \emph{open system}, when the simulation box
with gel freely exchanges ions with a big amount of aqueous solution
of certain constant salinity. 
\item And in \emph{closed system}, when the gel is in equilibrium with a
finite volume of aqueous solution.
\end{enumerate}
The compression of the gel in the \emph{open} system does not affect the surrounding salinity, thus the density of ions outside the gel remains constant $\cs =$ $\cna=$ $\ccl=$ $\text{Const}$ (\reffig{fig: open}). 
On the contrary, what remains constant in the \emph{closed} system, is the total number of $\na$ and $\cl$ ions, that are contained in the gel and in the external volume, \ie in the volume $\Vbox$ (\reffig{fig: closed}).

Let $\nna$ and $\ncl$ be the density of the corresponding ion in the compression volume, $\Vbox$ averaged over this volume
\begin{equation}
n_i = \left(N_i\gel + \cs\cdot(\Vbox - \Vgel)\right) / \Vbox
\end{equation}
where $ i \in \{\na, \cl\}$,  $N_i$ is the amount of the corresponding ions in gel.
Then in closed system $\nna$ and $\ncl$ remain constant, whereas the salinity of supernate, $\cs$, may change with the change of the gel volume.

@@

The similar scheme we use to simulate the gel in closed system. 
The procedure represents a chain of subsequent MD and MC simulation, but now the molecular dynamics is runing in two simulation boxes simultaneously.
The first box of the volume $\Vgel$ contains the gel with ions, and the second box, $\Vout$--- contains only ions. 
The total volume of both boxes is kept constant, $\Vbox=V_{\mathrm{gel}}+V_{\mathrm{out}}$, so the compression of the gel implies the decrease $\Vgel$ and increase $\Vout$.
These two boxes are represented in \reffig{fig:diamond}.

\paragraph{Algorithm.}
The thermodynamic equilibrium between the two volumes implies the exchange of the ion pairs between them, which is done by means of MC procedure in a way similar to described in \cite{Panagiotopoulos1988b}.
The simulation algorithm is following:
\begin{enumerate}
\item Simulate molecular dynamics in both boxes; collect the observables, for examples, values of pressure $P$, to an array; 
\item Perform Monte-Carlo procedure simulating ion pair exchange; collect the arrays of observables samples: number of ions in both boxes, $\nna\gel$,
$\ncl\gel$, $\nna\out$ and $\ncl\out$ ; 
\item Repeat the procedure until the desired length of sample arrays is reached.
\end{enumerate}
